### 2025-11-23 thesis/evaluation/index.tex
More evaluation content

\paragraph{System Configuration}

% TODO: Describe configuration parameters
% - Model loading settings
% - Batch sizes
% - Timeout configurations
% - Any other relevant parameters

@@

\paragraph{Spider Test Set}

The \textsc{Spider}

% TODO: Describe Spider benchmark
% - Dataset size: 10,181 questions, 5,693 SQL queries
% - 200 databases across 138 domains
% - Difficulty distribution (Easy/Medium/Hard/Extra Hard)
% - Why Spider was chosen
% - Train/dev/test split strategy

\paragraph{Bird Benchmark}
% TODO: Describe Bird benchmark
% - Dataset size: 12,751 pairs across 95 databases (33.4 GB)
% - 37 domains
% - Real-world characteristics (dirty data, external knowledge requirements)
% - Why Bird was chosen as complementary to Spider

\paragraph{Dataset Preprocessing}
% TODO: Describe any preprocessing steps
% - Database preparation
% - Schema extraction
% - Gold standard SQL formatting

@@

\paragraph{Execution Accuracy (EA)}
% TODO: Define execution accuracy
% - Definition: whether query results match ground truth
% - Calculation method
% - Why EA is the primary metric
% - Reference to natural-benchmark implementation

\paragraph{Exact Match (EM)}
% TODO: Define exact match
% - Definition: whether SQL query exactly matches ground truth
% - Calculation method
% - Limitations and interpretation

\paragraph{Performance Metrics}
% TODO: Define performance metrics
% - End-to-end latency (time from NL query to SQL result)
% - Per-component latency (σ, φ, π, ρ, ν timings)
% - Throughput (queries per second)
% - Resource utilization (GPU memory, CPU usage)

\paragraph{Statistical Significance}
% TODO: Describe statistical testing approach
% - Test selection (e.g., paired t-test, McNemar's test)
% - Significance threshold (p < 0.05)
% - Confidence intervals

@@

\paragraph{Internal Baselines}
% TODO: Describe ablated baselines
% - Zero-shot baseline (π only, no σ, φ, ρ, ν)
% - Partial system variants for ablation studies

\paragraph{State-of-the-Art Comparisons}
% TODO: List SOTA systems for comparison
% - GPT-4 results from literature
% - DIN-SQL, DAIL-SQL, C3 results
% - Other relevant open-source approaches
% - Note: Direct comparison constraints (different models, hardware)

\subsection{Benchmark Performance Analysis}

% TODO: Introduction paragraph presenting overall findings

@@

\paragraph{Overall Performance}
% TODO: Present overall EA and EM scores
% - Table: Overall Spider dev set results
% - Comparison to baseline
% - Comparison to SOTA from literature

\paragraph{Performance by Difficulty Level}
% TODO: Break down results by Spider difficulty classification
% - Table: EA/EM by Easy/Medium/Hard/Extra Hard
% - Analysis of performance degradation with difficulty
% - Discussion of which difficulty levels benefit most from system components

\paragraph{Performance by Database}
% TODO: Present per-database analysis
% - Identify databases where system excels
% - Identify databases where system struggles
% - Analyze characteristics that correlate with performance

\paragraph{Performance by SQL Complexity}
% TODO: Analyze results by SQL pattern complexity
% - Simple SELECT queries
% - JOIN operations
% - Nested subqueries
% - Aggregations and GROUP BY
% - Set operations (UNION, INTERSECT, EXCEPT)
% - Complex predicates

\paragraph{Learning Curve Analysis}
% TODO: Analyze impact of training example quantity
% - How does performance change with k-shot examples?
% - Diminishing returns analysis
% - Optimal k value determination

@@

\paragraph{Overall Performance}
% TODO: Present overall Bird results
% - Table: Overall Bird dev set results
% - Comparison to baseline
% - Comparison to SOTA from literature (GPT-4: 54.89%, human: 92.96%)

\paragraph{Domain-Specific Performance}
% TODO: Break down by Bird's 37 domains
% - Best-performing domains
% - Worst-performing domains
% - Domain characteristics that impact performance

\paragraph{Dirty Data Handling}
% TODO: Analyze performance on Bird's dirty data cases
% - How does system handle real-world data quality issues?
% - Error patterns related to data quality
% - Comparison to clean data performance

\paragraph{External Knowledge Requirements}
% TODO: Analyze cases requiring external knowledge
% - Identify queries needing domain knowledge
% - System performance on these cases
% - Limitations of pure schema-based approaches

@@

\paragraph{Open-Source vs Proprietary Models}
% TODO: Discuss open-source (OmniSQL 7B) vs proprietary (GPT-4) trade-offs
% - Performance gap analysis
% - Cost and accessibility advantages
% - Privacy and control benefits
% - Deployment feasibility differences

\paragraph{Qualitative Comparison}
% TODO: Beyond metrics, discuss systemic differences
% - System architecture differences
% - Component strategies (how others handle schema subsetting, refinement, etc.)
% - Integration and deployment characteristics

@@

\paragraph{Per-Component Timing Breakdown}
% TODO: Present timing for each pipeline component
% - Table: Average latency for σ, φ, π, ρ, ν
% - Distribution statistics (median, p50, p95, p99)
% - Identify bottlenecks

\paragraph{End-to-End Query Response Time}
% TODO: Present total pipeline latency
% - Distribution of total response times
% - Comparison to acceptable thresholds (what's "interactive"?)
% - Impact of query complexity on latency

\paragraph{Latency vs Accuracy Trade-offs}
% TODO: Analyze speed-accuracy Pareto frontier
% - Can we reduce latency by limiting refinement iterations?
% - Can we reduce latency by using fewer candidates for voting?
% - Configurable performance profiles (fast mode vs accurate mode)

\paragraph{Bottleneck Identification and Optimization Opportunities}
% TODO: Identify where time is spent
% - Model inference time (largest component?)
% - Embedding search time
% - Graph kernel computation time
% - Database execution time (for refinement)
% - Recommendations for optimization

>>>>

@@

\paragraph{Spider}

\textsc{Spider} is likely the most prevalently used benchmark for NL2SQL
systems in contemporary research. The test dataset is comprised of 10,181
questions and 5693 corresponding SQL queries spread across 200 databases
spanning 138 domains. Questions in \textsc{Spider} are categorized by
difficulty (easy, medium, hard, extra hard). Evaluations on \textsc{Spider}
refer to the test dataset.

\paragraph{Bird}

\textsc{Bird} is another widely used benchmark that is comprised of 12,751
question answer pairs spread across 95 databases in 37 domains. It aims to be
more representative of real world scenarios with external knowledge
requirements and is generally considered harder than \textsc{Spider}.

@@

\paragraph{Execution Accuracy (EA)}

\textsc{Execution Accuracy} (or EA) is the primary success metric of NL2SQL systems
as it represents the semantic accuracy of the SQL queries produced NL2SQL
systems. Execution accuracy is computed by determining whether the rows in the
candidate results are a permutation of the rows returned by ground truth
results.

For a query $q_g$ with candidate results $R_c$ and ground truth results $R_g$, 
where $R_c, R_g \subseteq \mathbb{V}^{n \times m}$ (sets of $n$ rows with $m$ columns 
of values $\mathbb{V}$), semantically accurate execution is defined as:

\begin{equation}
    semanticeq(R_c, R_g, q_g) = \mathbbm{1}\left[\exists \pi \in \Pi_m : 
    \begin{cases}
        R_c = \pi(R_g) & \text{if } ordered(q_g) \\
        R_c \equiv_{\text{multiset}} \pi(R_g) & \text{otherwise}
    \end{cases}
    \right]
\end{equation}

where $\Pi_m$ is the set of valid column permutations (those preserving column
value sets), $\pi(R_g)$ applies permutation $\pi$ to reorder columns in each
row of $R_g$, $\equiv_{\text{multiset}}$ denotes multiset equality (allowing
row reordering) and $ordered(q_g)$ is determining if the ground truth query
$q_g$ contains an \texttt{ORDER BY} clause. Row and value comparisons use
tolerance-based equality for floating-point values ($|\cdot| < 10^{-5}$). The
calculation for measuring \textsc{Natural}'s EA was ported from \textsc{Spider}
\citep{Spider}.

\paragraph{Exact Match (EM)}

% TODO: Define exact match
% - Definition: whether SQL query exactly matches ground truth
% - Calculation method
% - Limitations and interpretation

\textsc{Exact Match}..

\paragraph{Performance Metrics}

Measuring real world applicablility, the metric \textsc{Latency} is introduced 
for measuring the compound pipeline runtime, thus measuring NL query to final
SQL candidate.

% TODO:
% - End-to-end latency (time from NL query to SQL result)
% - Per-component latency (σ, φ, π, ρ, ν timings)
% - Throughput (queries per second)
% - Resource utilization (GPU memory, CPU usage)

@@

To determine the effect of approaches introduced by \textsc{Natural} two
a set of baselines is introduced. The baselines are split into two groups,
internal baselines (measuring the system without crucial components) and
external baselines (eg, State-of-the-Art system performance).

\paragraph{Internal Baselines}

Two primary internal baselines are introduced:

\begin{enumerate}
    \item \textbf{Model Baseline} — Using only the inference logic for
        zero-shot accuracy, aiming to represent \textsc{OmniSQL} performance as
        closely as possible.
    \item \textbf{Zero-Shot Baseline} — A full \textsc{Natural} pipeline but
        without any examples available for the in-context learning inference
        phase.
\end{enumerate}

These baselines aim to measure both the actual performance of the NL2SQL model
(\textsc{OmniSQL}) and the full \textsc{Natural} pipeline without in-context
learning.

\paragraph{State-of-the-Art Comparisons}

\textcolor{red}{X} external baselines are referenced from previous research:

% - GPT-4 results from literature
% - DIN-SQL, DAIL-SQL, C3 results
% - Other relevant open-source approaches
% - Note: Direct comparison constraints (different models, hardware)

\subsection{Benchmark Results}

@@

\paragraph{Overall Performance}

% TODO: Present overall EA and EM scores
% - Table: Overall Spider dev set results
% - Comparison to baseline
% - Comparison to SOTA from literature

\paragraph{Performance by Difficulty Level}

% TODO: Break down results by Spider difficulty classification
% - Table: EA/EM by Easy/Medium/Hard/Extra Hard
% - Analysis of performance degradation with difficulty
% - Discussion of which difficulty levels benefit most from system components

\paragraph{Performance by Database}

% TODO: Present per-database analysis
% - Identify databases where system excels
% - Identify databases where system struggles
% - Analyze characteristics that correlate with performance

\paragraph{Performance by SQL Complexity}

% TODO: Analyze results by SQL pattern complexity
% - Simple SELECT queries
% - JOIN operations
% - Nested subqueries
% - Aggregations and GROUP BY
% - Set operations (UNION, INTERSECT, EXCEPT)
% - Complex predicates

\paragraph{Learning Curve Analysis}

% TODO: Analyze impact of training example quantity
% - How does performance change with k-shot examples?
% - Diminishing returns analysis
% - Optimal k value determination

@@

\paragraph{Overall Performance}

% TODO: Present overall Bird results
% - Table: Overall Bird dev set results
% - Comparison to baseline
% - Comparison to SOTA from literature (GPT-4: 54.89%, human: 92.96%)

\paragraph{Domain-Specific Performance}

% TODO: Break down by Bird's 37 domains
% - Best-performing domains
% - Worst-performing domains
% - Domain characteristics that impact performance

\paragraph{Dirty Data Handling}

% TODO: Analyze performance on Bird's dirty data cases
% - How does system handle real-world data quality issues?
% - Error patterns related to data quality
% - Comparison to clean data performance

\paragraph{External Knowledge Requirements}

% TODO: Analyze cases requiring external knowledge
% - Identify queries needing domain knowledge
% - System performance on these cases
% - Limitations of pure schema-based approaches

@@

\paragraph{Open-Source vs Proprietary Models}

% TODO: Discuss open-source (OmniSQL 7B) vs proprietary (GPT-4) trade-offs
% - Performance gap analysis
% - Cost and accessibility advantages
% - Privacy and control benefits
% - Deployment feasibility differences

\paragraph{Qualitative Comparison}

% TODO: Beyond metrics, discuss systemic differences
% - System architecture differences
% - Component strategies (how others handle schema subsetting, refinement, etc.)
% - Integration and deployment characteristics

@@

\paragraph{Per-Component Timing Breakdown}

% TODO: Present timing for each pipeline component
% - Table: Average latency for σ, φ, π, ρ, ν
% - Distribution statistics (median, p50, p95, p99)
% - Identify bottlenecks

\paragraph{End-to-End Query Response Time}

% TODO: Present total pipeline latency
% - Distribution of total response times
% - Comparison to acceptable thresholds (what's "interactive"?)
% - Impact of query complexity on latency

\paragraph{Latency vs Accuracy Trade-offs}

% TODO: Analyze speed-accuracy Pareto frontier
% - Can we reduce latency by limiting refinement iterations?
% - Can we reduce latency by using fewer candidates for voting?
% - Configurable performance profiles (fast mode vs accurate mode)

\paragraph{Bottleneck Identification and Optimization Opportunities}

% TODO: Identify where time is spent
% - Model inference time (largest component?)
% - Embedding search time
% - Graph kernel computation time
% - Database execution time (for refinement)
% - Recommendations for optimization
